---
title: "Investigating Population Health and Economic Consequences of Weather Events in the United States"
author: "Thomas R Nudell"
date: "January 4, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r libraries}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

```{r working directory}
setwd("/home/trn/Dropbox/xxX/xxR/RRR/RepData_PeerAssignment2")
```


## Summary

This data analysis addresses the two following questions:

1. Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

[//]: # (Summary of analysis)

## Data Processing
[//]: # (Describe how data is loaded into R and processed for analysis, starting from Raw Data)

The data we will analize is a National Oceanic and Atmospheric Administration (NOAA) storm database. This data comes in the form of a comma seperated value (csv) file compressed with the bzip2 (bz2) algorithm.
We will name data file `StormData.csv.bz2`. 
```{r name data}
f.data <- "StormData.csv.bz2"
```

If the data doesn't already exists in the same directory as this script, we will need to download it from the course website. 

```{r download data}
if (!file.exists(f.data)) {
  f.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(f.url,f.data, method="curl")
}
```

We next read the data into memory. We will use `cache=TRUE` so we only have to read from the disk once during the course of analysis. 

```{r read data, cache=TRUE}
df <- read.csv(f.data)
```

### Preprocessing
Our data set contains information about 985 different types of events, listed in the `EVTYPE` variable, which can be seen by taking a peak at the unique entries. 
```{r unique events}
#print(unique(df$EVTYPE))
```

Looking at these unique entries also raises some questions about the cleanliness of this data. For example, `TSTM WIND` represents the same thing as `THUNDERSTORM WIND` which is the same as `THUNDERSTORM WINDS`, etc. In summary, our data is not clean and not tidy. However, because this project is not designed to test how well one can clean and tidy data, so I'll just walk through a few quick proceedures to tidy things up a bit. 

The documentation for this data, [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) only lists 48 different event types (see Table 1 on Page 6). I couldn't find a tablular form of this data so I copy/pasted Table 1 into a text file `EVTYPE.txt` and read it in using `read.delim` specifying a new line as the deliminator. 

```{r EVTYPE table}
EVTYPE <- read.delim('EVTYPE.txt', header=FALSE, sep = "\n")
```
The are "[C/Z/M]" characters at the end of the string that indicate what type of region the event was recorded in. Eventually it would be nice to give them their own column in our table. First we just extract the event types out. 

```{r EVTYPE$NAMES}
EVTYPE <- separate(EVTYPE,V1, into = "NAMES", sep="[C,Z,M]$")
EVTYPE$NAMES <- str_trim(EVTYPE$NAMES, side="both")
```

Next we do some convert the coded names to the closest true `EVTYPE`. If you know of a better way to do this, please let me know; I'm eager to learn. This first function will take an input string and return the closest element in `EVTYPE$NAMES`. It requires the `stringdist` package. 

```{r fixing EVTYPE in the df, cache=TRUE}
library("stringdist")
evtypeMatch <- function(X){
  as.character(EVTYPE$NAMES[which.min(stringdist(toupper(X),toupper(EVTYPE$NAMES)))])
} 

df <- mutate(df,EVTYPE = sapply(df$EVTYPE,evtypeMatch))
```


Next, we note that the values in `PROPDMG` and `CROPDMG` need to be adjusted by orders of magnitude according the entries in `PROPDMPEXP` and `CROPDMPEXP` repsectively. The valid values are “K” for thousands, “M” for millions, and “B” for billions. We will add a new variable which maps the alphabetical code to a numerical magnitude, then mutate the variables.  

```{r}
magnitude <- data.frame(ALPHA=c("K", "M","B"), MAG=c(1e3,1e6,1e9))
```


```{r}
test <-df %>% 
       select(EVTYPE, PROPDMG, CROPDMG, PROPDMGEXP, CROPDMGEXP) %>%
       mutate(CROPDMGEXP=toupper(CROPDMGEXP)) %>%
       mutate(PROPDMGEXP=toupper(PROPDMGEXP))

test <- left_join(x=test,y=magnitude,by=c('CROPDMGEXP'='ALPHA')) %>%
        mutate(CROPDMG=CROPDMG*MAG) %>%
        select(-MAG) %>%
        left_join(x=test,y=magnitude,by=c('PROPDMGEXP'='ALPHA')) %>%
        mutate(PROPDMG=PROPDMG*MAG) %>%
        select(-MAG)
head(test)
```



## Results 

[//]: # (Main results including 1--3 figures with descriptive captions)

### Population Health
We will measure population health impacts as the total number of `INJURIES` plus the total number of `FATALITIES` for a given event. TO clean things up a bit, we create a new dataframe with summarization of `INJURIES` plus `FATALITIES` by `EVTYPE`.
```{r population health}
health_df <- df %>% 
             select(EVTYPE,INJURIES,FATALITIES) %>%
             mutate(HEALTH_IMPACT = INJURIES + FATALITIES) %>%
             group_by(EVTYPE) %>%
             summarize(TOTAL_HEALTH = sum(HEALTH_IMPACT))
```

The first thing to notice is that most events have no health impact, so in order to get a better picture of impactful events, we filter out all event types with no health impacts. 

```{r filter non health impact events}
health_df <- filter(health_df, TOTAL_HEALTH != 0)
```

```{r plot evtypes}
g.health <- ggplot(health_df, aes(x=EVTYPE, y=TOTAL_HEALTH))
g.health + geom_point()
```


We can see in the plot that one event type has far and away the largest health impact, and that event type is "Tornado".
```{r}
print(health_df$EVTYPE[which.max(health_df$TOTAL_HEALTH)])
```

### Economic Consequences 

We measure economic consequences as the sum of property damage (`PROPDMG`), crop damage (`CROPDMG`). 

```{r economic consequences}
econ_df <-   df %>% 
             select(EVTYPE,PROPDMG,CROPDMG,PROPDMGEXP,CROPDMGEXP) %>%
             mutate(DMG = PROPDMG +CROPDMG) %>%
             group_by(EVTYPE) %>%
             summarize(TOTAL_DMG = sum(DMG)) %>%
             filter(TOTAL_DMG > 1e6)
```

Then we plot it. It looks like 
```{r econ plot}
g.econ <- ggplot(econ_df, aes(x = EVTYPE, y=TOTAL_DMG))
g.econ + geom_point()
```

