---
title: "Investigating Population Health and Economic Consequences of Weather Events in the United States"
author: "Thomas R Nudell"
date: "January 4, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This data analysis addresses the two following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

[//]: # (Summary of analysis)

## Data Processing
[//]: # (Describe how data is loaded into R and processed for analysis, starting from Raw Data)

The data we will analize is a National Oceanic and Atmospheric Administration (NOAA) storm database. This data comes in the form of a comma seperated value (csv) file compressed with the bzip2 (bz2) algorithm.
We will name data file `StormData.csv.bz2`. 
```{r}
f.data <- "StormData.csv.bz2"
```

If the data doesn't already exists in the same directory as this script, we will need to download it from the course website. Note that the `knitr` package automatically temporarilly changes the working directory to the directory of this script when executing code chunks, so there is no need to call `setwd()`. 

```{r}
if (!file.exists(f.data)) {
  f.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(f.url,f.data, method="curl")
}
```

We next read the data into memory. We will use `cache=TRUE` so we only have to read from the disk once during the course of analysis. 

```{r, cache=TRUE}
df <- read.csv(f.data)
```



## Results 

[//]: # (Main results including 1--3 figures with descriptive captions)